---
title: "the odin-monty toolkit"
author: "Marc Baguelin, Ed Knock, Rich FitzJohn"
date: "14-02-2025"
format: revealjs
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## odin - the beginnings

-   **odin** created to integrate ODEs (e.g. for compartmental models) in R with a domain-specific language (DSL)
-   later difference (discrete-time) equations
-   Simplifies the transition from mathematical model to efficient C++ code handled in R

## Pre-Pandemic

-   odin in use for various epidemiological and mathematical modelling tasks\
-   Provided a consistent DSL to keep modelling workflow organized and less error-prone

## Pandemic collaboration

-   **mcstate** and **dust** emerged during the COVID-19 pandemic\
-   Collaborative work by the UK real-time modelling & software engineers teams at Imperial College
-   Tailored to handle and fit complex state-space models at scale (e.g. Knock et al. 2021)

## Challenges addressed by dust and mcstate

-   Need for **efficient** (marginal) likelihood computation for large stochastic models
-   Exploit parallelism and particle filters
-   Built on top of odin’s approach, adding support for parameter inference and particle filtering

## Rethinking the design

-   Post-pandemic, recognized areas for improvement:
    -   More sophisticated data flows (better architecture for inference)\
    -   Support for probabilistic modeling extensions with an odin-like DSL
    -   Modular philosophy to handle dynamical (odin) and statistical (monty) models together

## Birth of odin2 and dust2 (Rich)

-\> @Rich for most people odin2 and dust2 are the same than odin and dust: can you summarise how the change make the whole toolkit more efficient? the parallelisation of parameter+stochastic draw groups is one I see and better memory usage between "fixed data" and changing parameters but feel free to put what you think

## Introducing monty

-   **monty**: A framework to handle statistical models (defined by density)\
-   Aims to offers multiple sampling solutions
-   Modular approach so that odin and monty "collaborate" but remain independent

## The odin-monty vision

-   Provide a **modern modeling ecosystem**:
    1.  Powerful DSL (odin2+monty)\
    2.  Streamlined simulation & filtering engine (dust2)\
    3.  User-friendly statistical integration platform (monty)\
-   Empower researchers to build robust models quickly, accelerating scientific insights

## Resources

-   We will provide a demonstration
-   You can find more in [our online odin-monty book](https://mrc-ide.github.io/odin-monty/)

## 2025 Roadmap {.smaller}

-   Some missing features from mcstate/dust1
    -   multi-stage filters
    -   restarting filters
    -   GPU-accelerated fitting
    -   SMC\^2 and IF (perhaps)
-   Some almost-ready features
    -   Parallel tempering (exploit parallelism, cope with multiple modes)
    -   Automatic differentiation to be extended to almost all odin models
    -   NUTS and HMC algorithms

## Odin-Monty Workshop {.smaller}
::: columns

:::: {.column width="60%"}
📅 **24–25 March 2025**  
📍 **White City, Imperial College London**  

✅ **Hybrid participation available**  
✅ **Travel grants available**  
✅ **Lightning talk slots!**  

🔗 **Register now [here]("https://docs.google.com/forms/d/e/1FAIpQLSePERaKXjMMYK0re1mh--2nNM5PCUZaRrd-SL_zSqAYY-pk5A/viewform")!**

⚠️ **Deadline: 19/02 – Don't miss out!** 


::::

:::: {.column width="40%"}
```{r, echo=FALSE}
#| fig-width: 8
code <- qrcode::qr_code("https://docs.google.com/forms/d/e/1FAIpQLSePERaKXjMMYK0re1mh--2nNM5PCUZaRrd-SL_zSqAYY-pk5A/viewform")
plot(code)
```
::::

:::

